The Impact of Accretion Disk Winds on the Optical Spectra of Cataclysmic Variables

### 2015-01-15 draft.tex
a few more changes discussion and runs for figs

\begin{abstract}
Many high-state non-magnetic cataclysmic variables (CVs) exhibit
blue-shifted absorption or P-Cygni profiles in ultraviolet
(UV) resonance lines, such as C~{\sc iv}~1550\AA. These features
imply the existence of powerful accretion disk winds in
CVs.
% and considerable effort has gone into modelling UV line formation
% in such outflows. By contrast, much less attention has been paid to
% the potential impact of these disk winds on the {\em optical} spectra
% of CVs. 
Here, we use an improved version of our Monte Carlo ionization
and radiative transfer code to investigate whether disk wind models that
produce realistic UV line profiles are also likely to generate
observationally significant recombination line and continuum emission
in the {\em optical} waveband.   
In particular, we test whether outflows may 
be responsible for the single-peaked emission line profiles often seen
in high-state CVs and for the weakness of the Balmer absorption edge
(relative to simple models of optically thick accretion disks).
We find that a standard disk wind model that is successful in
reproducing the UV spectra of CVs also leaves a noticeable imprint on
the optical spectrum, particularly for systems viewed at high
inclination. The strongest optical wind-formed recombinations lines are \ha\ and
\heiiopt\, although the outflow also produces emission in the Ly-$\alpha$
and \heiiuv\ UV lines. However, all of the optical lines produced by
this default wind model are double-peaked, and the effect of the wind
on the Balmer jump is modest. 
We demonstrate that a higher-density outflow model 
displays reasonable verisimilitude with the optical spectrum of RW Trianguli,
producing all the observed lines. In addition, this model
produces recombination continuum sufficient to fill in the Balmer jump,
although it also predicts stronger-than-observed 
collisionally excited UV emission lines. 
Our revised model also significantly
alters the spectral shape of the continuum and displays
a strong inclination dependence in its emission properties. 
No single-peaked emission is seen, although we observe a 
narrowing of the double-peaked emission lines from the base of the wind. 
We suggest that a non-homogenous, `clumpy' wind may provide a possible
solution.
On the basis of these results, we suggest that winds can 
modify, and perhaps even dominate, the line and continuum emission
from accreting systems.
% We carry out a limited exploration of
% the disk wind parameter space, focusing on those parameters that affect the
% density near the base of the wind. This shows that higher-density
% outflows can produce both single-peaked emission lines and a
% recombination continuum sufficient to fill in the Balmer
% jump, although such models also tend to predict stronger-than-observed 
% collisionally excited UV emission lines. Our main conclusion is that
% disk wind emission is likely to have a significant impact on the
% optical spectra of high-state CVs. 
% {\bf JM: This abstract is a few lines longer than I would like- 
% I think something punchier is required.}
\end{abstract}

@@

Murray \& Chiang (1996, 1997; hereafter referred to collectively as MC96)\nocite{MC96, MC97} 
have shown that the presence of disk winds may
offer a natural explanation for the single-peaked optical emission lines in
high-state CVs, since they can strongly affect the radiative transfer
of line photons. Strong support for a significant wind contribution to the
optical emission lines comes from observations of eclipsing
systems. There, the single-peaked lines are often only weakly
eclipsed, and a significant fraction of the line flux remains visible
even near mid-eclipse (REF). This points to line formation in a spatially
extended region, such as a disk wind (see figure~\ref{novalikes}).
%%(Honeycutt, Schlegel, \& Kaitchuck 1986; Dhillon \& Rutten 1995). 
Further evidence for a wind contribution to the optical lines comes
from isolated observations of P-Cygni-like line profiles even in optical
lines, such as \ha\ and He \textsc{i} $\lambda5876$ \citep{patterson1996, RN98, kafka2004}.

@@

Attempts to fit the observed spectral energy distributions (SEDs) of
high-state CVs with such models have met with mixed success. In
particular, the SEDs predicted by most stellar/disk atmosphere models 
are too blue in the UV \citep{wade1988,long1991,long1994}(Other refs here?) and exhibit
stronger-than-observed Balmer jumps in absorption 
\citep{wade1984,haug1987,ladous1989b,knigge1998}. One possible
explanation for these problems is that these models fail to capture
all of the relevant physics. Indeed, it has been argued that a
self-consistent treatment can 
produce better agreement with observational data (e.g. Shaviv et
al. 1991;  but also more recent work by Idan et al. 2010).
\nocite{idanshaviv2010} \nocite{shaviv1991}
However, an alternative explanation, suggested by \cite{KLWB98}, 
is that recombination continuum emission from the base of the 
disk wind might fill in the disk's
Balmer absorption edge and flatten the UV spectrum.

@@

We adopt the kinematic disk wind model developed by Schlosman \&
Vitello (1993; hereafter SV93). A schematic of this model is shown in
figure~\ref{cartoon}. In this parametrization, a smooth, biconical
disk wind emanates from the accretion disk between radii $r_{min}$ and 
$r_{max}$. The covering fraction of the outflow is also controlled by the
inner and outer opening angles of the wind, $\theta_{min}$ and
$\theta_{max}$, and the launch angle of the other streamlines is given
by 
\begin{equation}
\theta(r_0) = \theta_{min} + (\theta_{max} - \theta_{min}) \left(\frac{r_0 - r_{min}}{r_{max} - r_{min}} \right)^{\gamma},
\label{theta}
\end{equation}
where $r_0$ is the launch radius of the streamline.

@@

Figure~\ref{spec} shows the corresponding optical spectra produced for
the benchmark model, and these do exhibit some emission lines
associated with H and He. 
We see a general trend from absorption lines to emission lines 
with increasing inclination, as one might expect from our wind
geometry. This trend is consistent with observations, as can be  seen
in Figure~1. However, it is clear that this particular model
does not produce all the lines seen in observations of high-state CVs.
The higher order Balmer series lines too weak
to overcome the intrinsic absorption from the disk atmosphere, and the wind 
fails to produce any emission at intermediate inclinations, 
which contrasts with (for example) the observed behaviour of IX Vel
\citep{beuermann1990}.

Our models do not exhibit P-Cygni profiles in the optical lines.
This is perhaps not surprising. LK02 and SV93 originally designed such models
to reproduce the UV line profiles. Thus, most of the wind
has an ionization parameter of $\log U \sim 2$ (see figure 6).
This means Hydrogen and Helium are fully ionized throughout 
much of the wind and is successful in producing recombination features.
However, it leads to absorbing columns in \ha\ of $N_H\sim ??$ - far too
low to produce noticeable blue shifted absorption. 
While we suspect that the winds that exhibit such profiles must 
lie in a different ionization regime, requiring 
significantly different disk SEDs or mass-loss rates to a 
`standard' high-state CV. However, we note it is difficult for this
idea to exist within a line-driven wind framework.

@@

Two other features are worth noting in the optical
spectrum. First, the collisionally excited Ca~{\sc ii} H\&K 
emission lines at 3969\AA\ and (especially) 3934\AA\ 
become quite prominent in our densest models. Second, our model predicts a detectable
He~\textsc{ii} recombination line at 3202\AA. This is the Helium
equivalent of Paschen~$\beta$ and should be expected in all systems that
feature a strong He~\textsc{ii}~4686\AA\ line (the Helium
equivalent of Paschen~$\alpha$). 
%%Although it has been observed (REF),
This line is somewhat unfamiliar observationally, because it 
lies bluewards of the atmospheric cut-off, but
also redwards of most ultraviolet spectra. 
% {\bf JM: we should we reference something here for the 3202 line. 
% I blieve CK found a paper which shows it in an IP? I'll try and find it.}

@@

There are two possible ways in which the model
could be altered to better reproduce single-peaked
lines as a result of radial velocity shear. First,
one could adopt a more equatorial wind geometry. 
This would increase the velocity gradient in the radial 
direction. However, CV winds are known to be moderately collimated.
This is demonstrated by the fact that blue-shifted resonance line 
absorption is seen in low inclination systems such as RW Sex \citep{greensteinoke1982},
but not in eclipsing systems \citep{cordovamason1985}.
Furthermore, the main reason for adopting this type of 
wind geometry is that it already successfully models the 
P-Cygni resonance lines seen in the UV (LK02, Noebauer et al. 2010),
We note that equatorial winds are expected in AGN and X-ray binaries 
\citep{ponti2012}, and that these winds may be more likely
to exhibit the required kinematics.
Second, one could adopt velocity laws with higher 
velocity gradients in the line emission region. 
However, this poses problems for a smooth biconical wind. 
We have already demonstrated that high densities are required 
to produce significant emission. In a smooth model, which
satisfies the mass continuity condition (see section 3.1), 
it is therefore difficult to reconcile 
these high densities with the high velocity gradients required.

We suggest that clumping
in CV winds provides a possible solution.
Non-homogeneity could allow higher density pockets 
of emission to exist with radial velocity shear still being high.
Line-driven winds are expected to be unstable (REFS),
and a number of implementations have been used in the stellar wind
community (REFS). We therefore hope that further investigation of
clumped winds could help solve the current `mystery' of single-peaked 
line emission in high inclination systems.

% Although \ha\ is single-peaked in our model, the majority of the
% lines are double-peaked, particularly  
% Figure~\ref{halpha} shows how the H$\alpha$
% % He~{\sc ii}~$\lambda 4686$ 
% line changes with the kinematics of the wind
% for an inclination of $80^\circ$. 
% % The width and strength of He~{\sc ii} 
% % clearly depend on both $R_v$ and $\alpha$, but the overall line profile 
% % shape is always double-peaked. This contrasts with the behaviour of 
% The line becomes single-peaked in our densest models. 
% The transition from double-peaked to single-peaked occurs when the 
% line becomes optically thick. This reproduces the behaviour predicted by fsingle-pe96, who
% showed that the poloidal velocity gradient in the wind can modify the
% shape of optically thick emission lines even in the near-disk regime,
% where the poloidal velocity itself is much smaller than the rotational
% velocity. This has the effect of suppressing the line wings and
% can therefore lead to the production of a single-peaked line profile
% shape.

\begin{figure}
\includegraphics[width=0.5\textwidth]{figures/mc.eps}
\caption{
\ha\ line profiles, normalised to 1, plotted in veliocity space 
for three models with varying kinematic 
properties, computed at an inclination of $80^\circ$.
The benchmark model and the improved optical
model described in section 6 are labeled as A and B respectively,
and a third model (X) which has an increased acceleration length of 
$R_V = 283.8~R_{WD}$, and $\alpha=4$ is also shown. 
The $x$-axis limits correspond to the Keplerian velocity at 
$4R_{WD}$, the inner edge of the wind.
We observe a narrowing of the lines
but this is not due to the radial velocity shear mechanism (see section 5.3
for a discussion).
% The parameter that has been altered from the benchmark model 
% is marked in each column and row. Models towards the bottom right
% are have more slowly accelerating winds and denser wind bases. 
% {\bf JM: Need to discuss the `single-peaked-ness' of line profiles
% in our telecon.}
}
\label{halpha}
\end{figure} %fullpage

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%          DISCUSSION
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \section{The Dependence of Wind Features on Outflow Parameters}
%  Here we could discuss single-peaked emission and trends with parameters.
% \label{modelb}

% The results of section~\ref{modela} show that wind models that reproduce
% the UV spectral features of CVs are likely to leave a clear imprint on
% the optical spectrum. It is therefore important to understand the
% sensitivity of optical (and UV) wind-formed features to the basic
% properties of the outflow. In particular, there are three key questions
% we would like to address: (i) can a modified wind model account 
% for all of the lines we see in optical spectra of CVs while preserving
% the UV behaviour;
% (ii) under what conditions can disk winds
% produce {\em single-peaked} Balmer emission lines (MC96); 
% (ii) can the wind-formed recombination continuum completely
% fill in the disk's Balmer absorption edge for reasonable outflow
% parameters?

% The amount of recombination and collisionally excited emission
% produced by a given ion, $i$, in a plasma of volume $V$ depends
% primarily on the emission measure, $EM$, defined as 
% \begin{equation}
% EM=\int^V_0 n_e n_i \,dV,
% \end{equation}
% where $n_i$ is the number density of the ion in question.    
% The emission reaching an observer at a given frequency also depends on
% the Sobolev optical depth it experiences, which depends on both the
% density and the velocity gradient projected along the line of
% sight. This can also significantly affect line profile shapes.

% The simplest way to simultaneously affect the density in the wind as
% well as the velocity gradients is by modifying the poloidal velocity
% law. We have therefore conducted a exploratory sensitivity study in
% which we focus on just two kinematic variables (section~\ref{kinematics}):

% \begin{itemize}
%  	\item the acceleration length, $R_v$, which controls the
%         distance over which the wind accelerates to $\frac{1}{2}v_{\infty}$;
%  	\item the acceleration exponent, $\alpha$, which controls the rate 
%  	at which the poloidal velocity changes near $R_v$.
% \end{itemize} 
% The general behaviour we might expect is that outflows with denser
% regions near the wind base -- i.e. winds with larger $R_{v}$ and/or
% larger $\alpha$ -- will produce stronger optical emission signatures. 
% However, this behaviour may be moderated by the effect of the increasing
% optical depth through this region.

% \subsection{Optical Wind Signatures}
% \label{modelb_opt}

% Figure~\ref{halpha} shows how overall spectrum
% changes with the kinematics of the wind for an inclination
% of $80^\circ$. The width and strength of He~{\sc ii} clearly depend on
% both $R_v$ and $\alpha$, but the overall line profile shape is always
% double-peaked. This contrasts with the behaviour of H$\alpha$, which
% becomes single-peaked in our densest models. The transition from
% double-peaked to single-peaked occurs when the line becomes optically
% thick. This reproduces the behaviour predicted by \cite{MC96}, who
% showed that the poloidal velocity gradient in the wind can modify the
% shape of optically thick emission lines even in the near-disk regime,
% where the poloidal velocity itself is much smaller than the rotational
% velocity. This has the effect of suppressing the line wings and
% can therefore lead to the production of a single-peaked line profile
% shape.

% Figure~\ref{halpha} shows how the H$\alpha$ and 
% He~{\sc ii}~$\lambda 
% 4686$ lines change with the kinematics of the wind for an inclination
% of $80^\circ$. The width and strength of He~{\sc ii} clearly depend on
% both $R_v$ and $\alpha$, but the overall line profile shape is always
% double-peaked. This contrasts with the behaviour of H$\alpha$, which
% becomes single-peaked in our densest models. The transition from
% double-peaked to single-peaked occurs when the line becomes optically
% thick. This reproduces the behaviour predicted by \cite{MC96}, who
% showed that the poloidal velocity gradient in the wind can modify the
% shape of optically thick emission lines even in the near-disk regime,
% where the poloidal velocity itself is much smaller than the rotational
% velocity. This has the effect of suppressing the line wings and
% can therefore lead to the production of a single-peaked line profile
% shape.

% In order to produce single-peaked line emission in the majority of
% lines at high inclinations, even higher opacity is needed in the
% line wings. This difficulty in generating single-peaked line emission
% from a rotating disk wind was also encountered by SDL05, who found
% that an outflow geometry similar to that adopted here could not 
% reproduce the observed single peaked profiles in YSOs. For the
% parameters we have tested, the mechanism proposed by \cite{MC96} does
% produce single-peaked profiles, but only for most optically thick
% emission lines. In order for the same mechanism to affect the majority
% of lines, even higher wind densities would be required.

% Figure~\ref{jump} shows the behaviour of the Balmer jump region as a
% function of the kinematic wind parameters. Here, we clearly see the
% critical effect of the emission measure, with the densest winds
% filling in the disk's Balmer absorption edge most effectively.
% \footnote{Note that the apparent absorption feature 
% just redward of the Balmer jump in these models is artificial. It is
% caused by residual line blanketing in the stellar atmospheres, which
% our models cannot fill in since they employ a 20-level Hydrogen atom.}
% This offers a potential solution to a long-standing problem in
% understanding the spectral energy distribution of CVs, as
% originally suggested by \cite{KLWB98}.

% Two other features are worth noting in this part of the
% spectrum. First, the collisionally excited Ca~{\sc ii} H\&K 
% emission lines at 3969\AA\ 
% and (especially) 3934\AA\ become quite prominent in our densest
% models. Second, {\em all} of our models predict a detectable
% He~\textsc{ii} recombination line at 3202\AA. This is the Helium
% equivalent of Paschen~$\beta$ and should be expected in all systems that
% feature a strong He~\textsc{ii}~4686\AA\ line (the Helium
% equivalent of Paschen~$\alpha$). This line is unfamiliar observationally,
% but only because it lies bluewards of the atmospheric cut-off, but
% also redwards of most ultraviolet spectra. 
% {\bf JM: we should we reference something here for the 3202 line. 
% I blieve CK found a paper which shows it in an IP? I'll try and find it.}

% \begin{figure} %fullpage
% \centering
% \includegraphics[width=1.0\textwidth]{figures/3by3_grid_balmer.eps}  
% %%\includegraphics[width=0.5\textwidth]{figures/test.eps}
% \caption{
% The evolution of the Balmer jump with varying kinematic 
% properties, computed for an inclination of $80^\circ$. 
% The panels are organised in the same way as figure~\ref{halpha}.
% {\bf JM: Mark lines here?}
% % The panels are ordered in the same way as figure~\ref{halpha}. In general, increasing densities
% % lead to a Balmer jump in emission, although as in the case of H$\alpha$ there is a transition
% % when the wind becomes a neutral absorbing column.
% }
% \label{jump}
% \end{figure} %fullpage

% \subsection{Ultraviolet Wind Signatures}
% \label{modelb_uv}

% The dependence of the ultraviolet C~{\sc iv} and He~{\sc ii} lines on
% the same kinematic wind parameters is illustrated in figures~\ref{uvlines}
% and~\ref{uvlines}. As before, we focus on the predicted behaviour at $i =
% 80^\circ$. In our benchmark model, the C~{\sc iv} emission
% line seen at high inclinations is almost entirely produced via
% resonance scattering. The angle-averaged flux integrated over the
% entire line profile - accounting correctly for the blue-shifted
% absorption component seen at low inclinations - would be
% approximately zero. However, as we increase the density and extent of
% the slow-moving part of the outflow near the disk plane, collisionally
% excited emission starts to contribute significantly to the C~{\sc iv}
% line profiles. This explains the much stronger line-to-continuum
% contrast seen in the $\alpha = 1.5$, $R_v = 2\times 10^{11}$~cm model,
% for example. For any given inclination, the trend is not necessarily
% monotonic, however, because the escape probabilities from the
% line-forming region can be highly anisotropic.  
% {\bf CK: I *think* this is the basic reason, but (a) can you check
%   this, James; (b) is it enough to just point this out?; (c) do we
%   understand why CIV is single-peaked but HeII, for example, is double-peaked?
%   JM: There are still a few tests I want to do here- we can discuss in the telecon.
%   I believe He II v CIV shape is just due to CIV being produced in an extended region,
%   and generally scattered, whereas He II is produced close to the disk, and there's probably
%   also a line opacity effect.
%   I can check this.}

% The He~{\sc ii}~1640\AA recombination line is always double-peaked in
% our models, although the peak-to-peak separation decreases in the
% denser models. For double-peaked lines
% formed in rotating media -- in this case, the dense base of the wind
% -- the location of the peak(s) is roughly associated with the
% projected rotational velocity at the outer edge of the line-forming
% region. Thus the trend observed in He~{\sc ii} reflect the radial
% expansion of the line-forming region associated with an increase in
% the density at the base of the wind. The same trend is actually also
% seen in the optical He~{\sc ii}~4686\AA line. The strength of these
% features relative to their adjacent continua also track
% each other approximately. The reason for the non-monotonic behaviour
% in the line strengths as a function of the kinematic wind parameters
% is the same as for the C~{\sc iv} line.
% {\bf CK: again, can we check this is correct?}

% \begin{figure} %fullpage
% \mbox{
% % \subfigure{\includegraphics[width=0.33\textwidth]{figures/3by3_grid_lyman.eps}}  
% % \quad
% \subfigure{\includegraphics[width=0.5\textwidth]{figures/3by3_grid_uv.eps}}
% \quad
% \subfigure{\includegraphics[width=0.5\textwidth]{figures/3by3_grid_1640.eps}}  
% }
% %%\includegraphics[width=0.5\textwidth]{figures/test.eps}
% \caption{
% Line profiles for \civ\ (left) and \heiiuv\  (right) with varying kinematic 
% properties, computed for an inclination of $80^\circ$. 
% The panels are organised in the same way as figure~\ref{halpha}.
% {\bf JM: I need to adjust the smoothing here so for all plots I smooth
% in terms of Wavelength rather than bins. Linked to single-peaked discussion. 
% Note also CK may want a Lyman alpha plot, which I might put here}
% % The panels are ordered in the same way as figure~\ref{halpha}. In general, increasing densities
% % lead to a Balmer jump in emission, although as in the case of H$\alpha$ there is a transition
% % when the wind becomes a neutral absorbing column.
% }
% \label{uvlines}
% \end{figure}

@@

We have investigated whether a disk wind model designed to reproduce
the UV spectra of high-state CVs would also have a significant effect
on the optical spectra of these systems. We find that this is indeed
the case. In particular, the model wind produces H and He
recombination lines, as well as a recombination continuum blueward of
the Balmer edge. We do not produce P-Cygni profiles
in the optical H and He lines, 
which are seen in a small fraction of CV optical spectra.
Possible reasons for this are briefly discussed in section 
4.2.

@@

Although we note that limb darkening effects on disk atmosphere opacities
would affect the Balmer jump (see section 5.2). In section 5.3, we demonstrate that
although single-peaked emission can be formed in our densest models, 
it is not due to the radial velocity shear mechanism proposed by MC96.
The emission in our models which produce single-peaked lines comes from
approximately $10^{11}$~cm above the disk plane, which is not supported
by eclipse-mapping observations (REFS). We suggest
that non-homogenous, 
`clumpy' line-driven winds may offer a potential solution.

>>>>

\begin{abstract}
Many high-state non-magnetic cataclysmic variables (CVs) exhibit
blue-shifted absorption or P-Cygni profiles in ultraviolet
(UV) resonance lines, such as C~{\sc iv}~1550\AA. These features
imply the existence of powerful accretion disk winds in
CVs.
% and considerable effort has gone into modelling UV line formation
% in such outflows. By contrast, much less attention has been paid to
% the potential impact of these disk winds on the {\em optical} spectra
% of CVs. 
Here, we use an improved version of our Monte Carlo ionization
and radiative transfer code to investigate whether disk wind models that
produce realistic UV line profiles are also likely to generate
observationally significant recombination line and continuum emission
in the {\em optical} waveband.   
In particular, we test whether outflows may 
be responsible for the single-peaked emission line profiles often seen
in high-state CVs and for the weakness of the Balmer absorption edge
(relative to simple models of optically thick accretion disks).
We find that a standard disk wind model that is successful in
reproducing the UV spectra of CVs also leaves a noticeable imprint on
the optical spectrum, particularly for systems viewed at high
inclination. The strongest optical wind-formed recombinations lines are \ha\ and
\heiiopt\, although the outflow also produces emission in the Ly-$\alpha$
and \heiiuv\ UV lines. However, all of the optical lines produced by
this default wind model are double-peaked, and the effect of the wind
on the Balmer jump is modest. 
We demonstrate that a higher-density outflow model 
displays reasonable verisimilitude with the optical spectrum of RW Trianguli,
producing all the observed lines. In addition, this model
produces recombination continuum sufficient to fill in the Balmer jump,
although it also predicts stronger-than-observed 
collisionally excited UV emission lines. 
Our revised model also significantly
alters the spectral shape of the continuum and displays
a strong inclination dependence in its emission properties. 
No single-peaked emission is seen, although we observe a 
narrowing of the double-peaked emission lines from the base of the wind. 
We suggest that a nonhomogenous, `clumpy' wind may provide a possible
solution.
On the basis of these results, we suggest that winds can 
modify, and perhaps even dominate, the line and continuum emission
from accreting systems.
% We carry out a limited exploration of
% the disk wind parameter space, focusing on those parameters that affect the
% density near the base of the wind. This shows that higher-density
% outflows can produce both single-peaked emission lines and a
% recombination continuum sufficient to fill in the Balmer
% jump, although such models also tend to predict stronger-than-observed 
% collisionally excited UV emission lines. Our main conclusion is that
% disk wind emission is likely to have a significant impact on the
% optical spectra of high-state CVs. 
% {\bf JM: This abstract is a few lines longer than I would like- 
% I think something punchier is required.}
\end{abstract}

@@

Murray \& Chiang (1996, 1997; hereafter referred to collectively as MC96)\nocite{MC96, MC97} 
have shown that the presence of disk winds may
offer a natural explanation for the single-peaked optical emission lines in
high-state CVs, since they can strongly affect the radiative transfer
of line photons. Strong support for a significant wind contribution to the
optical emission lines comes from observations of eclipsing
systems. There, the single-peaked lines are often only weakly
eclipsed, and a significant fraction of the line flux remains visible
even near mid-eclipse \citep[e.g.][]{groot2004}. This points to line formation in a spatially
extended region, such as a disk wind (see figure~\ref{novalikes}).
%%(Honeycutt, Schlegel, \& Kaitchuck 1986; Dhillon \& Rutten 1995). 
Further evidence for a wind contribution to the optical lines comes
from isolated observations of P-Cygni-like line profiles even in optical
lines, such as \ha\ and He \textsc{i} $\lambda5876$ \citep{patterson1996, RN98, kafka2004}.

@@

Attempts to fit the observed spectral energy distributions (SEDs) of
high-state CVs with such models have met with mixed success. In
particular, the SEDs predicted by most stellar/disk atmosphere models 
are too blue in the UV \citep{wade1988,long1991,long1994} and exhibit
stronger-than-observed Balmer jumps in absorption 
\citep{wade1984,haug1987,ladous1989b,knigge1998}. One possible
explanation for these problems is that these models fail to capture
all of the relevant physics. Indeed, it has been argued that a
self-consistent treatment can 
produce better agreement with observational data (e.g. Shaviv et
al. 1991;  but also more recent work by Idan et al. 2010).
\nocite{idanshaviv2010} \nocite{shaviv1991}
However, an alternative explanation, suggested by \cite{KLWB98}, 
is that recombination continuum emission from the base of the 
disk wind might fill in the disk's
Balmer absorption edge and flatten the UV spectrum.

@@

We adopt the kinematic disk wind model developed by SV93. 
A schematic of this model is shown in
figure~\ref{cartoon}. In this parametrization, a smooth, biconical
disk wind emanates from the accretion disk between radii $r_{min}$ and 
$r_{max}$. The covering fraction of the outflow is also controlled by the
inner and outer opening angles of the wind, $\theta_{min}$ and
$\theta_{max}$, and the launch angle of the other streamlines is given
by 
\begin{equation}
\theta(r_0) = \theta_{min} + (\theta_{max} - \theta_{min}) \left(\frac{r_0 - r_{min}}{r_{max} - r_{min}} \right)^{\gamma},
\label{theta}
\end{equation}
where $r_0$ is the launch radius of the streamline.

@@

Figure~\ref{spec} shows the corresponding optical spectra produced for
the benchmark model, and these do exhibit some emission lines
associated with H and He. 
We see a general trend from absorption lines to emission lines 
with increasing inclination, as one might expect from our wind
geometry. This trend is consistent with observations, as can be  seen
in Figure~1. However, it is clear that this particular model
does not produce all the lines seen in observations of high-state CVs.
The higher order Balmer series lines are too weak
to overcome the intrinsic absorption from the disk atmosphere, and the wind 
fails to produce any emission at intermediate inclinations, 
which contrasts with (for example) the observed behaviour of IX Vel
\citep{beuermann1990}.

@@

Two other features are worth noting in the optical
spectrum. First, the collisionally excited Ca~{\sc ii} H\&K 
emission line at 3934\AA\ 
becomes quite prominent in our densest models. Second, our model predicts a detectable
He~\textsc{ii} recombination line at 3202\AA. This is the Helium
equivalent of Paschen~$\beta$ and should be expected in all systems that
feature a strong He~\textsc{ii}~4686\AA\ line (the Helium
equivalent of Paschen~$\alpha$). 
%%Although it has been observed (REF),
This line is somewhat unfamiliar observationally, because it 
lies bluewards of the atmospheric cut-off, but
also redwards of most ultraviolet spectra. 
% {\bf JM: we should we reference something here for the 3202 line. 
% I blieve CK found a paper which shows it in an IP? I'll try and find it.}

Our models do not exhibit P-Cygni profiles in the optical lines.
This is perhaps not surprising. LK02 and SV93 originally designed such models
to reproduce the UV line profiles. Thus, most of the wind
has an ionization parameter of $\log U \sim 2$ (see figure 6).
This means Hydrogen and Helium are fully ionized throughout 
much of the wind and is successful in producing recombination features.
However, it leads to absorbing columns in H of $N_H\sim ??$ - far too
low to produce noticeable blue shifted absorption. 
We therefore suspect that the winds that exhibit such profiles must 
possess a higher degree of ionization stratification.
This could manifest as a clumpy flow, in which ionization changes
due to small scale density fluctuations, or a stratification on acceleration 
length scales such as the kind shown by figure 6.
This idea is supported by the fact that \cite{prinja2000}
found dwarf nova BZ Cam's outflow to be unsteady and highly mass-loaded in outburst,
by observing the UV resonance lines in outburst with the {\sl Hubble Space Telescope}.
This system has exhibited P-Cygni profiles in He~\textsc{i}~$\lambda5876{\rm \AA}$
and \ha\ when in a high-state \cite{patterson1996,RN98}.

@@

There are two possible ways in which the model
could be altered to better reproduce single-peaked
lines as a result of radial velocity shear. First,
one could adopt a more equatorial wind geometry. 
This would increase the velocity gradient in the radial 
direction. However, CV winds are known to be moderately collimated.
This is demonstrated by the fact that blue-shifted resonance line 
absorption is seen in low inclination systems such as RW Sex \citep{greensteinoke1982},
but not in eclipsing systems \citep{cordovamason1985}.
Furthermore, the main reason for adopting this type of 
wind geometry is that it already successfully models the 
P-Cygni resonance lines seen in the UV (LK02, Noebauer et al. 2010),
Equatorial winds are expected in AGN and X-ray binaries 
\citep{weymann1991, ponti2012}, so these winds may be more likely
to possess the required kinematics.
Second, one could adopt velocity laws with higher 
velocity gradients in the line emission region. 
However, this poses problems for a smooth biconical wind. 
We have already demonstrated that high densities are required 
to produce significant emission. In a smooth model, which
satisfies the mass continuity condition (see section 3.1), 
it is therefore difficult to reconcile 
these high densities with the high velocity gradients required.

We suggest that clumping
in CV winds provides a possible solution.
Inhomogeneity could allow higher density pockets 
of emission to exist while still possessing a high radial velocity shear.
Theory of line-driven winds predicts an unstable flow
\citep{macgregor1979,owockirybicki1984,owockirybicki1985},
and simulations of accretion disk winds find a nonhomogeneous 
density structure \citep{proga1998,pkdh2002}.
In addition, a number of implementations have been used in the stellar wind
community \citep[e.g.][]{hamannkoesterke1998,oskinova2007}. 
We therefore hope that further investigation of
`clumpy' winds could help solve the current mystery of single-peaked 
line emission in high inclination systems.

\begin{figure}
\includegraphics[width=0.5\textwidth]{figures/mc.eps}
\caption{
\ha\ line profiles, normalised to 1, plotted in velocity space 
for three models with varying kinematic 
properties, computed at an inclination of $80^\circ$.
The benchmark model and the improved optical
model described in section 6 are labeled as A and B respectively,
and a third model (X) which has an increased acceleration length of 
$R_V = 283.8~R_{WD}$, and $\alpha=4$ is also shown. 
The $x$-axis limits correspond to the Keplerian velocity at 
$4R_{WD}$, the inner edge of the wind.
We observe a narrowing of the lines
but this is not due to radial velocity shear (see section 5.3
for a discussion).
}
\label{halpha}
\end{figure} %fullpage

@@

@@

We have investigated whether a disk wind model designed to reproduce
the UV spectra of high-state CVs would also have a significant effect
on the optical spectra of these systems. We find that this is indeed
the case. In particular, the model wind produces H and He
recombination lines, as well as a recombination continuum blueward of
the Balmer edge. We do not produce P-Cygni profiles
in the optical H and He lines, 
which are seen in a small fraction of CV optical spectra.
Possible reasons for this are briefly discussed in section 
5.2.

@@

Although we note that limb darkening effects on disk atmosphere opacities
would affect the Balmer jump (see section 5.2). In section 5.3, we demonstrate that
although single-peaked emission can be formed in our densest models, 
it is not due to the radial velocity shear mechanism proposed by MC96.
The emission in our models which produce single-peaked lines comes from
approximately $10^{11}$~cm above the disk plane, which is not supported
by eclipse-mapping observations (REFS). We suggest
that nonhomogenous, `clumpy' line-driven winds may offer a potential solution.
